---
title: "F0_F1_Popgen"
author: "Samuel Gurr"
date: "2024-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# SET WORKING DIRECTORY 

knitr::opts_knit$set(root.dir = "C:/Users/samjg/Documents/Github_repositories/EAD-ASEB-Airradians_Popgen_OA/HPC_analysis/") # Sam's
#knitr::opts_knit$set(root.dir = "C:/Users/samuel.gurr/Documents/Github_repositories/EAD-ASEB-Airradians_Popgen_OA/HPC_analysis/") # Sam's

```

#### load packages

```{r load packages we need}
library(vcfR)
library(hierfstat)
library(pcadapt)
library(adegenet)
library(pegas)
# install.packages('vcf2ploidy')
# library(vcf2ploidy) # https://github.com/dandewaters/VCF2Ploidy # use interactive shiny script to covert to colony
# library(stackr) # devtools::install_github("thierrygosselin/stackr")
library(sequoia)
library(qqman)
library(dplyr)
library(tidyr)
# library(pegas)
# library(LDlinkR)
# library(ldsep) # need a Seqinfo object  Airradians not registered in package
# library(updog) # need a Seqinfo object  Airradians not registered in package
# library(VariantAnnotation) # need a Seqinfo object  Airradians not registered in package
library(RcppCNPy)
library(ggfortify)
library(tidyverse)
library(ggrepel)
```


## load files

* vcf.gz files 

  
```{r load vcf and bed files: all samples and broodstock only}

getwd()
path = "output/Popgen/angsd/all/"


# Call the file with all samples - filtered using vcftools on SEDNA (check out the readme for what was done)
all.vcf    <- read.vcfR(paste0(path,"vcf/all_final.vcf.gz"))
all.bed    <- read.pcadapt(paste0(path,"plink/all_final.bed"), type = "bed") # 3897 variants
all.strata <- read.csv(paste0(path,"strata/all_final_listIDs.csv"), sep =',', header=FALSE) %>% 
                                dplyr::rename(Individual = V1) %>% 
                                dplyr::mutate(
                                      Individual = gsub('*./','',Individual),
                                      Type = dplyr::case_when(grepl("-B", Individual) ~ "broodstock", TRUE ~ 'juvenile'),
                                      Gen = dplyr::case_when(grepl("F0", Individual) ~ "F0",
                                                             grepl("F1", Individual) ~ "F1",
                                                             grepl("F2", Individual) ~ "F2",
                                                             grepl("F3", Individual) ~ "F3",
                                                             TRUE ~ "F1"),
                                      Treatment = dplyr::case_when(
                                        grepl("F0", Individual) ~ "none",
                                        grepl("pH7\\.",Individual) ~ "High",
                                        grepl(c("pH75\\.|.201.|.203.|.204.|.251.|.253.|.254.|.301.|.303.|.304.|.351.|.352.|.353.|.354."), Individual) ~
                                        "Moderate",
                                        grepl(c("pH8|.101.|.103.|.104.|.153.|.154.|.155.|.3.|.4.|.5."), Individual) ~ 
                                        "Low")) %>% 
                                dplyr::mutate(Gen_Treatment = 
                                                dplyr::case_when(Gen == "F0" ~ "F0",
                                                                 Gen %in% c("F1","F2","F3") ~ paste0(Gen,'_',Treatment)))
all.vcf

# pcangsd outputs .cov file 
all_cov_quad <- as.matrix(read.table(paste0(path, "plink/pcangsd/all_final.cov"), header = F))

```

```{r load vcf and bed files: all samples chromosome split}

getwd()
path = "output/Popgen/angsd/all/"


# Call the file with all samples - filtered using vcftools on SEDNA (check out the readme for what was done)
chrom1.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084264.1.vcf.gz"))
chrom1.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom1/chrom_CM084264.1.bed"), type = "bed") # 3897 variants

chrom2.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084265.1.vcf.gz"))
chrom2.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom2/chrom_CM084265.1.bed"), type = "bed") # 3897 variants

chrom3.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084266.1.vcf.gz"))
chrom3.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom3/chrom_CM084266.1.bed"), type = "bed") # 3897 variants

chrom4.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084267.1.vcf.gz"))
chrom4.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom4/chrom_CM084267.1.bed"), type = "bed") # 3897 variants

chrom5.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084268.1.vcf.gz"))
chrom5.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom5/chrom_CM084268.1.bed"), type = "bed") # 3897 variants

chrom6.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084269.1.vcf.gz"))
chrom6.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom6/chrom_CM084269.1.bed"), type = "bed") # 3897 variants

chrom7.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084270.1.vcf.gz"))
chrom7.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom7/chrom_CM084270.1.bed"), type = "bed") # 3897 variants

chrom8.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084271.1.vcf.gz"))
chrom8.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom8/chrom_CM084271.1.bed"), type = "bed") # 3897 variants

chrom9.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084272.1.vcf.gz"))
chrom9.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom9/chrom_CM084272.1.bed"), type = "bed") # 3897 variants

chrom10.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084273.1.vcf.gz"))
chrom10.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom10/chrom_CM084273.1.bed"), type = "bed") # 3897 variants

chrom11.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084274.1.vcf.gz"))
chrom11.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom11/chrom_CM084274.1.bed"), type = "bed") # 3897 variants

chrom12.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084275.1.vcf.gz"))
chrom12.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom12/chrom_CM084275.1.bed"), type = "bed") # 3897 variants

chrom13.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084276.1.vcf.gz"))
chrom13.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom13/chrom_CM084276.1.bed"), type = "bed") # 3897 variants

chrom14.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084277.1.vcf.gz"))
chrom14.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom14/chrom_CM084277.1.bed"), type = "bed") # 3897 variants

chrom15.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084278.1.vcf.gz"))
chrom15.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom15/chrom_CM084278.1.bed"), type = "bed") # 3897 variants

chrom16.vcf    <- read.vcfR(paste0(path,"vcf/all_final_chrom_split/chrom_CM084279.1.vcf.gz"))
chrom16.bed    <- read.pcadapt(paste0(path,"plink/all_final_chrom_split/chrom16/chrom_CM084279.1.bed"), type = "bed") # 3897 variants



# pcangsd outputs .cov file 
chrom1_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom1/pcangsd/chrom1_pcangsd.cov"), header = F))
chrom2_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom2/pcangsd/chrom2_pcangsd.cov"), header = F))
chrom3_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom3/pcangsd/chrom3_pcangsd.cov"), header = F))
chrom4_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom4/pcangsd/chrom4_pcangsd.cov"), header = F))
chrom5_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom5/pcangsd/chrom5_pcangsd.cov"), header = F))
chrom6_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom6/pcangsd/chrom6_pcangsd.cov"), header = F))
chrom7_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom7/pcangsd/chrom7_pcangsd.cov"), header = F))
chrom8_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom8/pcangsd/chrom8_pcangsd.cov"), header = F))
chrom9_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom9/pcangsd/chrom9_pcangsd.cov"), header = F))
chrom10_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom10/pcangsd/chrom10_pcangsd.cov"), header = F))
chrom11_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom11/pcangsd/chrom11_pcangsd.cov"), header = F))
chrom12_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom12/pcangsd/chrom12_pcangsd.cov"), header = F))
chrom13_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom13/pcangsd/chrom13_pcangsd.cov"), header = F))
chrom14_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom14/pcangsd/chrom14_pcangsd.cov"), header = F))
chrom15_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom15/pcangsd/chrom15_pcangsd.cov"), header = F))
chrom16_cov_quad <- as.matrix(read.table(paste0(path, "plink/all_final_chrom_split/chrom16/pcangsd/chrom16_pcangsdls.cov"), header = F))

```

```{r load vcf and bed files: broodstock only}

getwd()
path = "output/Popgen/angsd/all/"

# all broodstock 
broodstock.vcf    <- read.vcfR(paste0(path,"vcf/broodstock_final.vcf.gz"))
broodstock.bed    <- read.pcadapt(paste0(path,"plink/broodstock_final.bed"), type = "bed") # 3897 variants
broodstock.strata <- read.csv(paste0(path,"strata/broodstock_listIDs.csv"), sep =',', header=FALSE)  %>% 
                                dplyr::rename(Individual = V1) %>% 
                                dplyr::mutate(Individual = gsub('*./','',Individual),
                                      Gen = dplyr::case_when(grepl("F0", Individual) ~ "F0",
                                                             grepl("F1", Individual) ~ "F1",
                                                             grepl("F2", Individual) ~ "F2",
                                                             grepl("F3", Individual) ~ "F3",
                                                             TRUE ~ "F1"),
                                      Treatment = dplyr::case_when(
                                        grepl("F0", Individual) ~ "none",
                                        grepl("pH7\\.",Individual) ~ "High",
                                        grepl("pH75\\.",Individual) ~ "Moderate",
                                        grepl("pH8\\.",Individual) ~ "Low")) %>% 
                                dplyr::mutate(Gen_Treatment = 
                                                dplyr::case_when(Gen == "F0" ~ "F0",
                                                                 Gen %in% c("F1","F2","F3") ~ paste0(Gen,'_',Treatment)))                            
# pcangsd outputs .cov file 
broodstock_cov_quad <- as.matrix(read.table(paste0(path, "plink/pcangsd/broodstock_final.cov"), header = F))
  
``` 


# Tidy the vcf

* Why / What does this do?

```{r tidy the vcf file}
# Tidy the VCFs

# all samples
all_tidy <- all.vcf %>%
  vcfR2tidy(verbose = FALSE)

# only broodstock
broodstock_tidy <- broodstock.vcf %>%
  vcfR2tidy(verbose = FALSE)
```

# Genlight object

* What is it?

* Why

```{r red the genlight obj}
# Read in the genlight
all_genlight <- all.vcf %>%
                vcfR2genlight()
```

# Keep only the mapped chromosomes 

```{r}

# example of the locus name CM084264.1_39013, these are chroosome positions 
# whereas those starting with JAYE are not
# Keep only loci mapped to chromosomes:
# only chromosome accession IDs
loc_chrom_tbl <- tibble(locus = all_genlight@loc.names) %>%
                    extract(locus, "chrom", "CM0(\\d+)", remove = FALSE) %>%
                    extract(locus, "pos", "_(\\d+)", remove = FALSE) %>%
                    mutate(chrom = as.integer(chrom),
                           pos = as.integer(pos)) 
loc_chrom_tbl <- dplyr::filter(loc_chrom_tbl, !grepl("JAYEEO",locus)) %>% 
                 dplyr::mutate(chrom_num = 
                             case_when(chrom %in% 84264 ~ 1,
                                       chrom %in% 84265 ~ 2,
                                       chrom %in% 84266 ~ 3,
                                       chrom %in% 84267 ~ 4,
                                       chrom %in% 84268 ~ 5,
                                       chrom %in% 84269 ~ 6,
                                       chrom %in% 84270 ~ 7,
                                       chrom %in% 84271 ~ 8,
                                       chrom %in% 84272 ~ 9,
                                       chrom %in% 84273 ~ 10,
                                       chrom %in% 84274 ~ 11,
                                       chrom %in% 84275 ~ 12,
                                       chrom %in% 84276 ~ 13,
                                       chrom %in% 84277 ~ 14,
                                       chrom %in% 84278 ~ 15,
                                       chrom %in% 84279 ~ 16
                                       ))
nrow(loc_chrom_tbl) # 3130 total SNP calls

chrom_loci <- loc_chrom_tbl %>%
                filter(chrom_num < 20) %>%
                pull(locus)

chroms <- loc_chrom_tbl %>%
                filter(chrom_num < 20) %>%
                mutate(chrom_name = substr(locus,1,10)) %>% 
                pull(chrom_name) %>%
                unique()

# other contigs starting with accession delimiter JAYEE
loc_othercontigs_tbl <- tibble(locus = all_genlight@loc.names) %>%
                            extract(locus, "othercontigs", "JAYEEO(\\d+)", remove = FALSE) %>%
                            extract(locus, "pos", "_(\\d+)", remove = FALSE) %>%
                            mutate(othercontigs = as.integer(othercontigs),
                                   pos = as.integer(pos))
loc_othercontigs_tbl <- dplyr::filter(loc_othercontigs_tbl, !grepl("CM0",locus))
nrow(loc_othercontigs_tbl) # 767 total SNP calls

```


```{r}
# Create the genotype matrix
all_gl_mat <- all_genlight %>%
                    as.matrix()

gl_chrom <- all_genlight[,chrom_loci]

```


## plot PCAngsd cov data

* protocol and script from https://github.com/aersoares81/PCAngsd-PCA-Plots-in-R?tab=readme-ov-file

```{r plot PCAngsd cov data}

# Now it will perform the PCA

# all data 
all_run.pca        <- eigen(all_cov_quad)

#chromosomes split data for all individuals
chrom1_run.pca         <- eigen(chrom1_cov_quad)
chrom2_run.pca         <- eigen(chrom2_cov_quad)
chrom3_run.pca         <- eigen(chrom3_cov_quad)
chrom4_run.pca         <- eigen(chrom4_cov_quad)
chrom5_run.pca         <- eigen(chrom5_cov_quad)
chrom6_run.pca         <- eigen(chrom6_cov_quad)
chrom7_run.pca         <- eigen(chrom7_cov_quad)
chrom8_run.pca         <- eigen(chrom8_cov_quad)
chrom9_run.pca         <- eigen(chrom9_cov_quad)
chrom10_run.pca        <- eigen(chrom10_cov_quad)
chrom11_run.pca        <- eigen(chrom11_cov_quad)
chrom12_run.pca        <- eigen(chrom12_cov_quad)
chrom13_run.pca        <- eigen(chrom13_cov_quad)
chrom14_run.pca        <- eigen(chrom14_cov_quad)
chrom15_run.pca        <- eigen(chrom15_cov_quad)
chrom16_run.pca        <- eigen(chrom16_cov_quad)

# broodstock data 
broodstock_run.pca <- eigen(broodstock_cov_quad)

# Here is the information from your population: location and name of the samples
# note you can use the all samples strata for the chromosome split - its the same number of individuals 
all_samplemat     <- as.matrix(all.strata)
all_sampleheaders <- c("sample_id", "Type", "Gen", "Treatment", "Gen_Treatment")
colnames(all_samplemat) <- all_sampleheaders


broodstock_samplemat     <- as.matrix(broodstock.strata)
broodstock_sampleheaders <- c("sample_id", "Gen", "Treatment", "Gen_Treatment")
colnames(broodstock_samplemat) <- broodstock_sampleheaders


# Extract the eigenvectors and turn them into a dataframe for plotting
all_eigenvectors = all_run.pca$vectors #extract eigenvectors 

chrom1_eigenvectors  = chrom1_run.pca$vectors #extract eigenvectors 
chrom2_eigenvectors  = chrom2_run.pca$vectors #extract eigenvectors 
chrom3_eigenvectors  = chrom3_run.pca$vectors #extract eigenvectors 
chrom4_eigenvectors  = chrom4_run.pca$vectors #extract eigenvectors 
chrom5_eigenvectors  = chrom5_run.pca$vectors #extract eigenvectors 
chrom6_eigenvectors  = chrom6_run.pca$vectors #extract eigenvectors 
chrom7_eigenvectors  = chrom7_run.pca$vectors #extract eigenvectors 
chrom8_eigenvectors  = chrom8_run.pca$vectors #extract eigenvectors 
chrom9_eigenvectors  = chrom9_run.pca$vectors #extract eigenvectors 
chrom10_eigenvectors = chrom10_run.pca$vectors #extract eigenvectors 
chrom11_eigenvectors = chrom11_run.pca$vectors #extract eigenvectors 
chrom12_eigenvectors = chrom12_run.pca$vectors #extract eigenvectors 
chrom13_eigenvectors = chrom13_run.pca$vectors #extract eigenvectors 
chrom14_eigenvectors = chrom14_run.pca$vectors #extract eigenvectors 
chrom15_eigenvectors = chrom15_run.pca$vectors #extract eigenvectors 
chrom16_eigenvectors = chrom16_run.pca$vectors #extract eigenvectors 

broodstock_eigenvectors = broodstock_run.pca$vectors #extract eigenvectors 

# Combining the information from the populations/samples
all_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(all_eigenvectors)))

chrom1_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom1_eigenvectors))) %>% dplyr::mutate(chrom_num = 1)
chrom2_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom2_eigenvectors))) %>% dplyr::mutate(chrom_num = 2)
chrom3_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom3_eigenvectors))) %>% dplyr::mutate(chrom_num = 3)
chrom4_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom4_eigenvectors))) %>% dplyr::mutate(chrom_num = 4)
chrom5_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom5_eigenvectors))) %>% dplyr::mutate(chrom_num = 5)
chrom6_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom6_eigenvectors))) %>% dplyr::mutate(chrom_num = 6)
chrom7_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom7_eigenvectors))) %>% dplyr::mutate(chrom_num = 7)
chrom8_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom8_eigenvectors))) %>% dplyr::mutate(chrom_num = 8)
chrom9_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom9_eigenvectors))) %>% dplyr::mutate(chrom_num = 9)
chrom10_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom10_eigenvectors))) %>% dplyr::mutate(chrom_num = 10)
chrom11_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom11_eigenvectors))) %>% dplyr::mutate(chrom_num = 11)
chrom12_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom12_eigenvectors))) %>% dplyr::mutate(chrom_num = 12)
chrom13_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom13_eigenvectors))) %>% dplyr::mutate(chrom_num = 13)
chrom14_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom14_eigenvectors))) %>% dplyr::mutate(chrom_num = 14)
chrom15_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom15_eigenvectors))) %>% dplyr::mutate(chrom_num = 15)
chrom16_pca.vectors = as_tibble(cbind(all_samplemat, data.frame(chrom16_eigenvectors))) %>% dplyr::mutate(chrom_num = 16)

chromMaster_pca.vector <- rbind(chrom1_pca.vectors,chrom2_pca.vectors,chrom3_pca.vectors,chrom4_pca.vectors,
                                chrom5_pca.vectors, chrom6_pca.vectors,chrom7_pca.vectors, chrom8_pca.vectors,
                                chrom9_pca.vectors, #chrom10_pca.vectors, 
                                chrom11_pca.vectors, chrom12_pca.vectors,
                                chrom13_pca.vectors, #chrom14_pca.vectors,
                                chrom15_pca.vectors, chrom16_pca.vectors)

broodstock_pca.vectors = as_tibble(cbind(broodstock_samplemat, data.frame(broodstock_eigenvectors)))

# You can edit this too
legend_title <- "Gen"

# This will create a "simple" PCA plot. It will have a legend with the location of the samples next to it.
# Adjust the title, of course.
# Please notice that I commented the RColorBrewer.
# Uncomment and use the palette your like the most if you want to change the colors.
# Don't forget to uncomment the library in in the beginning of the script too.
all_simple.pca = ggplot(data = all_pca.vectors, 
                        aes(x=X1, y=X2, 
                        color = Treatment, 
                        shape = Gen, 
                        label = sample_id)) + 
                    geom_point(size=4) + 
                    labs(title="Title goes here", x="PC1",y="PC2") + 
                    # I'm making the title in italics because it's usually the name of a species
                    theme(plot.title = element_text(face = "italic")) #+ scale_color_brewer(palette="Set1")
plot(all_simple.pca)






chromMaster_juvenilesLow.pca = ggplot(data = (chromMaster_pca.vector %>% 
                                            filter(Type %in% 'juvenile') %>% 
                                            filter(Treatment %in% c('none','Low'))), 
                        aes(x=X1, y=X2, 
                        color = Gen, 
                        shape = Gen, 
                        label = sample_id)) + 
                    scale_shape_manual(values = c(19, 19, 19)) +
                    scale_color_manual(values = c('grey70', # 1 open point, F1
                                                  'grey40', # 19 solid point, F2
                                                  'black')) + # 19 solid point, F3
                    geom_point(size=2) + 
                    stat_ellipse() +
                    labs(title="Juveniles: Low OA", x="PC1",y="PC2") + 
                    theme_bw() +
                    # I'm making the title in italics because it's usually the name of a species
                    theme(plot.title = element_text(face = "italic")) +#+ scale_color_brewer(palette="Set1")
                    facet_wrap(~chrom_num)
plot(chromMaster_juvenilesLow.pca)

chromMaster_juvenilesModerate.pca = ggplot(data = (chromMaster_pca.vector %>% 
                                            filter(Type %in% 'juvenile') %>% 
                                            filter(Treatment %in% c('none','Moderate'))), 
                        aes(x=X1, y=X2, 
                        color = Gen, 
                        shape = Gen, 
                        label = sample_id)) + 
                    scale_shape_manual(values = c(19, 19, 19)) +
                    scale_color_manual(values = c('grey70', # 1 open point, F1
                                                  'grey40', # 19 solid point, F2
                                                  'black')) + # 19 solid point, F3
                    geom_point(size=2) + 
                    stat_ellipse() + 
                    labs(title="Juveniles: Moderate OA", x="PC1",y="PC2") + 
                    theme_bw() +
                    # I'm making the title in italics because it's usually the name of a species
                    theme(plot.title = element_text(face = "italic")) +#+ scale_color_brewer(palette="Set1")
                    facet_wrap(~chrom_num)
plot(chromMaster_juvenilesModerate.pca)

chromMaster_juvenilesHigh.pca = ggplot(data = (chromMaster_pca.vector %>% 
                                            filter(Type %in% 'juvenile') %>% 
                                            filter(Treatment %in% c('none','High'))), 
                        aes(x=X1, y=X2, 
                        color = Gen, 
                        shape = Gen, 
                        label = sample_id)) + 
                    scale_shape_manual(values = c(19, 19, 19)) +
                    scale_color_manual(values = c('grey70', # 1 open point, F1
                                                  'grey40', # 19 solid point, F2
                                                  'black')) + # 19 solid point, F3
                    geom_point(size=2) + 
                    stat_ellipse() +
                    labs(title="Juveniles: High OA", x="PC1",y="PC2") + 
                    theme_bw() +
                    # I'm making the title in italics because it's usually the name of a species
                    theme(plot.title = element_text(face = "italic")) +#+ scale_color_brewer(palette="Set1")
                    facet_wrap(~chrom_num)
plot(chromMaster_juvenilesHigh.pca)


# output plots
pdf("../RAnalysis/Output/Popgen/PCA_chromsplit_juveniles_LowOA.pdf", width = 8, height = 8)
plot(chromMaster_juvenilesLow.pca)
dev.off()

pdf("../RAnalysis/Output/Popgen/PCA_chromsplit_juveniles_ModerateOA.pdf", width = 8, height = 8)
plot(chromMaster_juvenilesModerate.pca)
dev.off()

pdf("../RAnalysis/Output/Popgen/PCA_chromsplit_juveniles_HighOA.pdf", width = 8, height = 8)
plot(chromMaster_juvenilesHigh.pca)
dev.off()








# Broodstock plots
chromMaster_broodstockLow.pca = ggplot(data = (chromMaster_pca.vector %>% 
                                                 filter(Type %in% 'broodstock') %>% 
                                                 filter(Treatment %in% c('none','Low'))), 
                        aes(x=X1, y=X2, 
                        color = Gen, 
                        shape = Gen, 
                        label = sample_id)) + 
                    scale_shape_manual(values = c(19, 19, 19)) +
                    scale_color_manual(values = c('grey70', # 1 open point, F1
                                                  'grey40', # 19 solid point, F2
                                                  'black')) + # 19 solid point, F3 
                    geom_point(size=2) + 
                    labs(title="Broodstock: Low OA", x="PC1",y="PC2") + 
                    theme_bw() +
                    # I'm making the title in italics because it's usually the name of a species
                    theme(plot.title = element_text(face = "italic"))  +
                    facet_wrap(~chrom_num)
plot(chromMaster_broodstockLow.pca)


chromMaster_broodstockModerate.pca = ggplot(data = (chromMaster_pca.vector %>% 
                                                 filter(Type %in% 'broodstock') %>% 
                                                 filter(Treatment %in% c('none','Moderate'))), 
                        aes(x=X1, y=X2, 
                        color = Gen, 
                        shape = Gen, 
                        label = sample_id))  + 
                    scale_shape_manual(values = c(19, 19, 19)) +
                    scale_color_manual(values = c('grey70', # 1 open point, F1
                                                  'grey40', # 19 solid point, F2
                                                  'black')) + # 19 solid point, F3 
                    geom_point(size=2) + 
                    labs(title="Broodstock: Moderate OA", x="PC1",y="PC2") + 
                    theme_bw() +
                    # I'm making the title in italics because it's usually the name of a species
                    theme(plot.title = element_text(face = "italic"))  +
                    facet_wrap(~chrom_num)
plot(chromMaster_broodstockModerate.pca)


chromMaster_broodstockHigh.pca = ggplot(data = (chromMaster_pca.vector %>% 
                                                 filter(Type %in% 'broodstock') %>% 
                                                 filter(Treatment %in% c('none','High'))), 
                        aes(x=X1, y=X2, 
                        color = Gen, 
                        shape = Gen, 
                        label = sample_id))  + 
                    scale_shape_manual(values = c(19, 19, 19)) +
                    scale_color_manual(values = c('grey70', # 1 open point, F1
                                                  'grey40', # 19 solid point, F2
                                                  'black')) + # 19 solid point, F3 
                    geom_point(size=2) + 
                    labs(title="Broodstock: High OA", x="PC1",y="PC2") + 
                    theme_bw() +
                    # I'm making the title in italics because it's usually the name of a species
                    theme(plot.title = element_text(face = "italic"))  +
                    facet_wrap(~chrom_num)
plot(chromMaster_broodstockHigh.pca)






chromMaster_broodstockF0F1.pca = ggplot(data = (chromMaster_pca.vector %>% 
                                                 filter(Type %in% 'broodstock') %>% 
                                                 filter(Gen %in% c('F0','F1'))), 
                        aes(x=X1, y=X2, 
                        color = Treatment, 
                        shape = Gen, 
                        label = sample_id))  + 
                    scale_shape_manual(values = c(1, 19)) +
                    scale_color_manual(values = c('grey70', # 1 open point, F1
                                                  'grey40', # 19 solid point, F2
                                                  'black')) + # 19 solid point, F3 
                    geom_point(size=2) + 
                    labs(title="F0 & F1: All OA", x="PC1",y="PC2") + 
                    theme_bw() +
                    stat_ellipse() +
                    # I'm making the title in italics because it's usually the name of a species
                    theme(plot.title = element_text(face = "italic"))  +
                    facet_wrap(~chrom_num)
plot(chromMaster_broodstockF0F1.pca)


chromMaster_broodstockF0BF3J.pca = ggplot(data = (chromMaster_pca.vector %>% 
                                                 mutate(Gen_Type = paste0(Gen, "_", Type)) %>% 
                                                 filter(Gen_Type %in% c('F0_broodstock', 'F3_juvenile'))),
                        aes(x=X1, y=X2, 
                        color = Treatment, 
                        shape = Gen, 
                        label = sample_id))  + 
                    scale_shape_manual(values = c(1, 19)) +
                    scale_color_manual(values = c('black',
                                                  'grey70', # 1 open point, F1
                                                  'grey40', # 19 solid point, F2
                                                  'black')) + # 19 solid point, F3 
                    geom_point(size=2) + 
                    labs(title="F0 Broodstock & F3 Juveniles: All OA", x="PC1",y="PC2") + 
                    theme_bw() +
                    stat_ellipse() +
                    # I'm making the title in italics because it's usually the name of a species
                    theme(plot.title = element_text(face = "italic"))  +
                    facet_wrap(~chrom_num)
plot(chromMaster_broodstockF0BF3J.pca)




# This will create another PCA plot.
# In this version, every point (or as many as possible) will have the sample ID assigned to it.
# It's neat if you want to see where exactly each sample is placed.
labeled.pca = ggplot(data = pca.vectors, aes(x=X1, y=X2, colour = Gen, label = sample_id)) + 
  geom_point() + labs(title="Title goes here", x="PC1",y="PC2") + 
  # I'm making the title in italics because it's usually the name of a species
  theme(plot.title = element_text(face = "italic")) + 
  geom_label_repel(show.legend = FALSE) #+ scale_color_brewer(palette="Set1")
plot(labeled.pca)

# This will output a PDF file with your plot, adjust the size and dpi to your liking.
#It will be created in your basedir.
ggsave(filename = paste0(basedir, "simple_pca_plot.pdf"), 
  plot = simple.pca, width = 28, height = 24, units = "cm", dpi = 600) 

# This will output a PDF with plot that has all the labels.
# It will be created in your basedir.
ggsave(filename = paste0(basedir, "labeled_pca_plot.pdf"), 
  plot = labeled.pca, width = 28, height = 24, units = "cm", dpi = 600)

```

```{r}

```

```{r}

```

```{r}

```


## About parentage analysis 

the choice of the package will depend on the specifc reesarch questions and the nature of the study species

**Sequoia**

a comprehensive R package for muligenerational pedigree reconstruction
emplpys a fast heuristic hill-climbing algorithm to explore the liklihood surface
using SNP data and input sinply the birthyear of your genotyped individuals

key features include: 

* parentage assignment 
* sibship clustering
* grandparent assignment 
* andles any proportion of genotyped parents
* accounts for genotyping errors 
* does not require predefined lists of candidate parents 


* use adegent - genlight is a dataframe compatible with adegenet based on genome-wide SNP data

```{r sequoia mock data}

data(Ped_HSg5, LH_HSg5)
# Simulate some genotype data to use for this try-out.
Geno <- SimGeno(Ped = Ped_HSg5, nSnp = 200)
# Run sequoia with the genotype data, lifehistory data, and otherwise default values. It is often advisable to first only run parentage assignment, and check if the results are sensible and/or if any parameters need adjusting. Full pedigree reconstruction, including sibship clustering etc., is much more time consuming.
ParOUT <- sequoia(GenoM = Geno,
                  LifeHistData = LH_HSg5,
                  Module = 'par')
ParOUT$PedigreePar
```

```{r build genlight}
?sequoia
# Go from vcf file to a geno file for sequoia
# create a genlight object from vcf file
F0BF1J_longcontigs_genlight <- vcfR2genlight(F0B_F1JALL_intersect.vcf, n.cores = 4)
?GenoConvert
View(Geno_Sequoia)
Geno_Sequoia <- GenoConvert(
                    InData = F0BF1J_longcontigs_genlight,
                    # InFile = NULL,
                    InFormat = "seq",
                    # OutFile = NA,
                    OutFormat = "seq")
# convert to matrix
Geno_Sequoia                <-  (as.matrix(F0BF1J_longcontigs_genlight))

delim_add <- stringr::str_replace_all(rownames(Geno_Sequoia), "_", ".")
splt <- strsplit(sub(".", "_", delim_add),".", fixed = TRUE)
rownames(Geno_Sequoia) <- sapply(splt, "[", 3)
rownames(Geno_Sequoia) <- paste0('ID_',rownames(Geno_Sequoia))

LH_Sequoia <- F0BroodF1Juveniles_strata %>% 
                dplyr::mutate(BirthYear = dplyr::case_when(
                                          Gen %in% 'F0' ~ 1,
                                          Gen %in% 'F1' ~ 2)) %>% 
                dplyr::select(Individual, BirthYear) %>% 
                dplyr::rename(ID = Individual)

delim_add <- stringr::str_replace_all(LH_Sequoia$ID, "_", ".")
splt <- strsplit(sub(".", "_", delim_add),".", fixed = TRUE)
LH_Sequoia$ID <- sapply(splt, "[", 3)

F0_parentage <- read.csv('output/lcWGS/angsd/F0_Broodstock/parentage.csv') %>% 
                    dplyr::mutate(Sample.ID = sub('.*adapter_trim.', '', Sample.ID)) %>% 
                    dplyr::rename(ID = Sample.ID, Sex = key)

# merge with parentage to assign Sex for squoia
# also change ID witha prexis so it can run
LH_Sequoia_2 <- merge(LH_Sequoia, F0_parentage,by="ID", all=T) %>% 
                      dplyr::mutate(
                        ID = paste0('ID_',ID),
                        Sex = dplyr::case_when(
                          Sex %in% 'Mo' ~ '1',
                          Sex %in% 'Pa' ~ '4',
                          Sex %in% 'Fa' ~ '2',
                          TRUE ~ Sex)) %>% 
                      dplyr::mutate(Sex = as.numeric(Sex))

# ?sequoia

ParOUT <-sequoia(GenoM = Geno_Sequoia,
                 LifeHistData = LH_Sequoia_2,
                 Module="par")
ParOUT$LifeHist

SeqOUT <-sequoia(GenoM= Geno_Sequoia,
                 SeqList =ParOUT,
                 Module="ped")

```





# Screeplot of bed file

-   what to look for?   the 'elbow' of this plot infers the number of descriptive principle components

-  assign the K value and rerun as '.res'

- Manhattan plot - regions involved in adaptation OR low recombination (high Linkage disequilibrium) affecting ascertainment of population structure,
when working with whole genome data we need to perform SNP thinning to acocutn for LD

  - What does LD look like on a Manhatttan plot? A regoin on the x axis SNPs with the same Y value (-log10 pvalues), appears 
  as a bar (many) or pair of points close/touching on the same Y-axis plane


```{r screeplot}

# INTERSECT DATA :::::::::::::::::::::::::::::::::::::::::::::::::::::::
## data: F0B_F1JALL_intersect.bed

## F0 brood with all F1 juvneiles
all.diagnostic      <- pcadapt(all.bed, K = 20)
ggpubr::ggarrange(
  plot(all.diagnostic, option = "screeplot"),
  plot(all.diagnostic, option = "manhattan"))

all.res             <- pcadapt(all.bed, K = 6)
ggpubr::ggarrange(
  plot(all.res, option = "screeplot"),
  plot(all.res, option = "manhattan"))
 
```



### Is there LD in my data, and is it problematic? (Evaluating loadings)

* To evaluate if LD might be an issue for your dataset... display the loadings (contributions of each SNP to the PC)
  
  - what do i look for? If the loadings are clustered in a single or several genomic regions.

```{r is there LD}

par(mfrow = c(2, 2))
for (i in 1:4) 
  plot(all.res$loadings[, i], pch = 19, cex = .3, ylab = paste0("Loadings PC", i))
```

### Linkage disequilibrium thinning using ```pcadapt```

- ```size``` = window size (default 200 SNPs), window radius of which SNPs are lumped 

- ```thr``` = r^2 threshold (default 0.1, correlation threshold 


* Lets write a function to **optimize** ```LD.cliumping```

```{r clump4me}

clump4me <- function(data, LD.thr, LD.K, LD.minMaf, outputfilename) {
  
        # For loop ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::;;; #
        outputfilename      <- data.frame()  # start dataframe 
        loop.table          <- data.frame(matrix(nrow = 1, ncol = 5)) # create dataframe, columns for the number of sampling days
        colnames(loop.table)<-c('output_PC1', 'input_MinMaf', 'input_Window', 'input_Threshold', 'input_NumPCAs') # names for comuns in the for loop
        LD.window_loop      <- seq(100,50100, by=1000) # run the loop for CPM from 1 - 10 (counts per million) 
        LD.window_loop      <- data.frame(LD.window_loop)
        
        
        for (i in 1:nrow(LD.window_loop)) {
          
          LD.window <- LD.window_loop[i,1]
          res       <- pcadapt(input = data, 
                         K = LD.K, 
                         LD.clumping = list(size = LD.window, 
                                            thr = LD.thr),
                         min.maf= LD.minMaf)  
        
          loop.table$input_MinMaf    = LD.minMaf
          loop.table$input_Window    = LD.window
          loop.table$input_Threshold = LD.thr
          loop.table$input_NumPCAs   = LD.K
          loop.table$output_PC1      = abs(mean(res$scores[,1]))
        
            
          # df             <- data.frame(loop.table) # name dataframe for this singl e row
          outputfilename <- rbind(outputfilename,loop.table) #bind to a cumulative list dataframe
          #print(df_total) # print to monitor progress
        }
        
        return(outputfilename)
}
```

```{r run clump4me choose settings and run}

library(ggpubr)

# INTERSECT DATA :::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

# *  run the bam file - sliding window 50 -1000 at settings
test <- as.data.frame(clump4me(data         = all.bed,
                             LD.thr         = 0.1, # r squares
                             LD.K           = 4, # number of PCAs
                             LD.minMaf      = 0.05, # minumum allele freq threshold (5% - =0.05, etc)
                             outputfilename = all_linkage))
plot(output_PC1~input_Window , data=test) # window of 250


# run with window size 5000, no rationale just to try it
all_LDdiagnostics <- pcadapt(all.bed, 
                             K = 20, 
                             LD.clumping = 
                               list(size = 5000, thr = 0.1),
                             min.maf= 0.05)
plot(all_LDdiagnostics, option = "screeplot") # Lets try and use K = 7 after SNP thinning 

all_LDres <- pcadapt(all.bed, 
                             K = 9, 
                             LD.clumping = 
                               list(size = 5000, thr = 0.1),
                             min.maf= 0.05)
ggarrange( (plot(all.res, option = "manhattan")),
           (plot(all_LDres, option = "manhattan")),nrow=2)

```


```{r F0B_F1JuvMODERATE qqman manhattan plots}


all_qqman           <- as.data.frame(matrix(nrow=nrow(all_tidy$fix), ncol=4)) 
colnames(all_qqman) <- c('SNP','CHR', 'BP', 'P')
all_qqman           <- all_qqman[-c(1:4)] %>% 
                                    dplyr::mutate(
                                      SNP = paste0(all_tidy$fix$CHROM, '_',all_tidy$fix$POS),
                                      CHR = as.numeric(extract_numeric(all_tidy$fix$CHROM)),
                                      BP  = all_tidy$fix$POS,
                                      P   = all_LDres$pvalues
                                    ) %>% na.omit() %>% 
                                     dplyr::mutate(CHR = 
                                             case_when(CHR %in% 84264.1 ~ 1,
                                                       CHR %in% 84265.1 ~ 2,
                                                       CHR %in% 84266.1 ~ 3,
                                                       CHR %in% 84267.1 ~ 4,
                                                       CHR %in% 84268.1 ~ 5,
                                                       CHR %in% 84269.1 ~ 6,
                                                       CHR %in% 84270.1 ~ 7,
                                                       CHR %in% 84271.1 ~ 8,
                                                       CHR %in% 84272.1 ~ 9,
                                                       CHR %in% 84273.1 ~ 10,
                                                       CHR %in% 84274.1 ~ 11,
                                                       CHR %in% 84275.1 ~ 12,
                                                       CHR %in% 84276.1 ~ 13,
                                                       CHR %in% 84277.1 ~ 14,
                                                       CHR %in% 84278.1 ~ 15,
                                                       CHR %in% 84279.1 ~ 16
                                                       ))

all_qqman %>% dplyr::filter(CHR < 90 & -log(all_qqman$P) > -log(0.001))



manhattan( (all_qqman %>% dplyr::filter(CHR < 90 & -log(all_qqman$P))),
            annotatePval = 0.001)


library(tidyr)
library(dplyr)
don <- all_qqman %>% 
  
  # Compute chromosome size
  group_by(CHR) %>% 
  summarise(chr_len=max(BP)) %>% 
  
  # Calculate cumulative position of each chromosome
  mutate(tot=cumsum(chr_len)-chr_len) %>%
  select(-chr_len) %>%
  
  # Add this info to the initial dataset
  left_join(all_qqman, ., by=c("CHR"="CHR")) %>%
  
  # Add a cumulative position of each SNP
  arrange(CHR, BP) %>%
  mutate( BPcum=BP+tot)

axisdf = don %>%
  group_by(CHR) %>%
  summarize(center=( max(BPcum) + min(BPcum) ) / 2 )

ggplot(don, aes(x=BPcum, y=-log10(P))) +
    
    # Show all points
    geom_point( aes(color=as.factor(CHR)), alpha=0.8, size=1.3) +
    scale_color_manual(values = rep(c("grey", "skyblue"), length(unique(don$CHR)) )) +
    
    # custom X axis:
    scale_x_continuous( label = axisdf$CHR, breaks= axisdf$center ) +
    scale_y_continuous(expand = c(0, 0) ) +     # remove space between plot area and x axis
  
    # Custom the theme:
    theme_bw() +
    theme( 
      legend.position="none",
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    )




```

```{r}
library(ggpubr)
# INTERSECT DATA :::::::::::::::::::::::::::::::::::::::::::::::::::::::
par(mfrow = c(1, 2))
for (i in 1:2)
  plot(F0B_F1JuvAll_intersect.LDres$loadings[, i], pch = 19, cex = .3, ylab = paste0("Loadings PC", i))

plot(F0B_F1JuvAll_intersect.LDres, option = "scores", pop = F0BroodF1Juveniles_strata$Gen)
plot(F0B_F1JuvAll_intersect.LDres, option = "scores", pop = F0BroodF1Juveniles_strata$Gen_Treatment)
plot(F0B_F1JuvAll_intersect.LDres , option = "manhattan")

# MERGED DATA ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

## all
ggarrange( (plot(F0B_F1JuvAll_merged.LDres, option = "scores", pop = F0BroodF1Juveniles_strata$Gen_Treatment)),
           (plot(F0B_F1JuvAll_merged.LDres , option = "manhattan")),
         nrow = 2)

## pH 8 LOW
ggarrange( (plot(F0B_F1JLOW_merged.LDres, option = "scores", pop = F0F1_strata_pH8$Gen)),
           (plot(F0B_F1JLOW_merged.LDres , option = "manhattan")),
         nrow = 2)

## pH 75 MODERATE
ggarrange( (plot(F0B_F1JMODERATE_merged.LDres, option = "scores", pop = F0F1_strata_pH75$Gen)),
           (plot(F0B_F1JMODERATE_merged.LDres , option = "manhattan")),
         nrow = 2)


```

This might look normal, but you’ll notice that two of the populations are tightly grouped around PC1. We should check too make sure this pattern isn’t being driven by a linkage in the genome. To do this, we can look at the loading scores of the PCs. Loading scores show how much a particular SNP factors into a PC.

```{r View PCAs 1 through 4}

# pH 8
par(mfrow = c(2, 2))
for (i in 1:4) {
  plot(pH8_res$loadings[, i], pch = 19, cex = .3, ylab = paste0("Loadings PC", i))
}


# pH 75 - this one looks wierd..
par(mfrow = c(2, 2))
for (i in 1:4) {
  plot(pH75_res$loadings[, i], pch = 19, cex = .3, ylab = paste0("Loadings PC", i))
}
```

# Linkage disequilibrium

-   Linkage Disequilibrium can affect ascertainment of population structure (Abdellaoui et al. 2013). Users analyzing dense data such as SNP Array data should account for LD in their PCAs and PCA-based genome-scans.

-   PCAdapt has a built in feature that allows us to clump SNPs by LD, reducing the data to a more informative subset.

```{r address the LD baby!}
library(ggpubr)


# LD_clumping: 
#   Default is NULL and doesn't use any SNP thinning. If you want to use SNP thinning, provide a named list with parameters $size and $thr which corresponds respectively to the window radius and the squared correlation threshold. A good default value would be list(size = 500, thr = 0.1)
  


# INTERSECT DATA :::::::::::::::::::::::::::::::::::::::::::::::::::::::
## data: F0B_F1JuvAll_intersect.bed,

F0B_F1JALL_intersect_LD <- pcadapt(F0B_F1JALL_intersect.bed, K = 10, LD.clumping = list(size = 500, thr = 0.1))
plot(F0B_F1JALL_intersect_LD, option = "screeplot") # looks like 2 PCAs
plot(F0B_F1JALL_intersect_LD, option = "scores", pop = F0BroodF1Juveniles_strata$Gen) # 
plot(F0B_F1JALL_intersect_LD, option = "scores", pop = F0BroodF1Juveniles_strata$Gen_Treatment) # 


# MERGED DATA ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
## data: F0B_F1JuvAll_merge.bed, F0 Broodstock w/ F1 Juveniles ALL
F0B_F1JALL_merged_LD <- pcadapt(F0B_F1JALL_merged.bed, K = 10, LD.clumping = list(size = 500, thr = 0.1))
plot(F0B_F1JALL_merged_LD, option = "screeplot") # looks like 2 PCAs
plot(F0B_F1JALL_merged_LD, option = "scores", pop = F0BroodF1Juveniles_strata$Gen) # F2 is wonky AF
plot(F0B_F1JALL_merged_LD, option = "scores", pop = F0BroodF1Juveniles_strata$Gen_Treatment) # F2 is wonky AF


## * F0 Broodstock w/ F1 Juveniles LOW
F0B_F1JLOW_merge_LD <- pcadapt(F0B_F1JLOW_merged.bed, K = 10, LD.clumping = list(size = 500, thr = 0.1))
plot(F0B_F1JLOW_merge_LD, option = "screeplot") # looks like 6
plot(F0B_F1JLOW_merge_LD, option = "scores", pop = F0F1_strata_pH8$Gen) #


## * F0 Broodstock w/ F1 Juveniles MODERATE
F0B_F1JMODERATE_merge_LD <- pcadapt(F0B_F1JMODERATE_merged.bed, K = 10, LD.clumping = list(size = 1000, thr = 0.2))
plot(F0B_F1JMODERATE_merge_LD, option = "screeplot") # looks like 2.5? PCAs
plot(F0B_F1JMODERATE_merge_LD, option = "scores", pop = F0F1_strata_pH75$Gen) #




# cxompare some plots to see!
library(ggpubr)

ggarrange(
  
      (plot(F0B_F1JuvAll_merge.res, option = "scores", 
            pop = F0BroodF1Juveniles_strata$Gen_Treatment)),
      
      (plot(F0B_F1JALL_merged_LD, option = "scores", 
            pop = F0BroodF1Juveniles_strata$Gen_Treatment))
        )


ggarrange(
  
      (plot(F0B_F1JuvLOW_merge.res, option = "scores", 
            pop = F0F1_strata_pH8$Gen)),
      
      (plot(F0B_F1JLOW_merge_LD, option = "scores", 
            pop = F0F1_strata_pH8$Gen))
        )
```


-   Interestingly, we see that compensating for LD, the four groups are more distinct, but not by treatment

-   what other attributes could be responsible for this separation?

    -   we had 4 replicates per treatment (A, B, C, D), however the buckets were randomly distributed at random and rotated. Perhaps there was a tank position during a critical stage (on the seawater table) that caused this separation? Unforuntatley we do not have replicate info for the broodstock, as these were pooled from each replicate and positions on the spawning table.

    -   Look at the F2 Juveniles (here we have A-D information) to see if this arises also


## Build genid objects

```{r}
# create genind object from vcf file - use the LD object 

# INTERSECT DATA ::::::::::::::::::::::::::::::::::::::::::::::::::::::::::
## data: F0B_F1JALL_intersect.vcf
F0B_F1JALL_intersect.genind       <- vcfR2genind(F0B_F1JALL_intersect.vcf)
F0B_F1JALL_intersect.vcf_FILTERED <- F0B_F1JALL_intersect.vcf[!is.na(F0B_F1JALL_intersect_LD$loadings[,1]),]
F0B_F1JALL_intersect.genind_LD    <- vcfR2genind(F0B_F1JALL_intersect.vcf_FILTERED)


strata(F0B_F1JALL_intersect.genind) <- F0BroodF1Juveniles_strata
setPop(F0B_F1JALL_intersect.genind) <- ~Gen_Treatment

strata(F0B_F1JALL_intersect.genind_LD) <- F0BroodF1Juveniles_strata
setPop(F0B_F1JALL_intersect.genind_LD) <- ~Gen_Treatment


F0B_F1JALL_intersect_tab    <- tab(F0B_F1JALL_intersect.genind, freq = TRUE, NA.method = "mean")
F0B_F1JALL_intersect_pca1   <- dudi.pca(F0B_F1JALL_intersect_tab, scale = FALSE, scannf = FALSE, nf = 3)
barplot(F0B_F1JALL_intersect_pca1$eig[1:25], 
        main = "PCA eigenvalues", 
        col = heat.colors(25))
s.class(F0B_F1JALL_intersect_pca1$li, pop(F0B_F1JALL_intersect.genind))


F0B_F1JALL_intersect.genind_LD_tab  <- tab(F0B_F1JALL_intersect.genind_LD, freq = TRUE, NA.method = "mean")
F0B_F1JALL_intersect.genind_LD_pca1 <- dudi.pca(F0B_F1JALL_intersect.genind_LD_tab, scale = FALSE, scannf = FALSE, nf = 3)
barplot(F0B_F1JALL_intersect.genind_LD_pca1$eig[1:25], 
        main = "PCA eigenvalues", 
        col = heat.colors(25))
s.class(F0B_F1JALL_intersect.genind_LD_pca1$li, pop(F0B_F1JALL_intersect.genind_LD))



# MERGED DATA :::::::::::::::::::::::::::::::::::::::::::::::::::::::
## data: F0B_F1JALL_merged.vcf, F0B_F1JLOW_merged.vcf, F0B_F1JMODERATE_merged.vcf

## * F0 Broodstock w/ F1 Juveniles ALL
F0B_F1JALL_merged.genind       <- vcfR2genind(F0B_F1JALL_merged.vcf) # 81,325
F0B_F1JALL_merged.vcf_FILTERED <- F0B_F1JALL_merged.vcf[!is.na(F0B_F1JALL_merged_LD$loadings[,1]),]
F0B_F1JALL_merged.genind_LD    <- vcfR2genind(F0B_F1JALL_merged.vcf_FILTERED) # 18,405 


strata(F0B_F1JALL_merged.genind) <- F0BroodF1Juveniles_strata
setPop(F0B_F1JALL_merged.genind) <- ~Gen_Treatment

strata(F0B_F1JALL_merged.genind_LD) <- F0BroodF1Juveniles_strata
setPop(F0B_F1JALL_merged.genind_LD) <- ~Gen_Treatment


F0B_F1JALL_merged_tab    <- tab(F0B_F1JALL_merged.genind, freq = TRUE, NA.method = "mean")
F0B_F1JALL_merged_pca1   <- dudi.pca(F0B_F1JALL_merged_tab, scale = FALSE, scannf = FALSE, nf = 3)
barplot(F0B_F1JALL_merged_pca1$eig[1:25], 
        main = "PCA eigenvalues", 
        col = heat.colors(25))
s.class(F0B_F1JALL_merged_pca1$li, pop(F0B_F1JALL_merged.genind))


F0B_F1JALL_merged.genind_LD_tab  <- tab(F0B_F1JALL_merged.genind_LD, freq = TRUE, NA.method = "mean")
F0B_F1JALL_merged.genind_LD_pca1 <- dudi.pca(F0B_F1JALL_merged.genind_LD_tab, scale = FALSE, scannf = FALSE, nf = 3)
barplot(F0B_F1JALL_merged.genind_LD_pca1$eig[1:25], 
        main = "PCA eigenvalues", 
        col = heat.colors(25))
s.class(F0B_F1JALL_merged.genind_LD_pca1$li, pop(F0B_F1JALL_merged.genind_LD))












## * F0 Broodstock w/ F1 Juveniles LOW
F0B_F1JLOW_merged.genind         <- vcfR2genind(F0B_F1JLOW_merged.vcf)

## * F0 Broodstock w/ F1 Juveniles MODERATE
F0B_F1JMODERATE_merged.genind    <- vcfR2genind(F0B_F1JMODERATE_merged.vcf)















pH8_genind               <- vcfR2genind(F1F2_pH8_vcf)
F1F2_pH8_vcf_ld_filtered <- F1F2_pH8_vcf[!is.na(pH8_res_LD$loadings[,1]),]
pH8_genind_ld_filtered   <- vcfR2genind(F1F2_pH8_vcf_ld_filtered)

pH75_genind               <- vcfR2genind(F1F2_pH75_vcf)
F1F2_pH75_vcf_ld_filtered <- F1F2_pH75_vcf[!is.na(pH75_res_LD$loadings[,1]),]
pH75_genind_ld_filtered   <- vcfR2genind(F1F2_pH75_vcf_ld_filtered)

# assign metadata to these loci 

strata(pH8_genind) <- F1_F2_pH8_strata_df
setPop(pH8_genind) <- ~Gen

strata(pH8_genind_ld_filtered) <- F1_F2_pH8_strata_df
setPop(pH8_genind_ld_filtered) <- ~Gen




strata(pH75_genind) <- F1_F2_pH75_strata_df
setPop(pH75_genind) <- ~Gen

strata(pH75_genind_ld_filtered) <- F1_F2_pH75_strata_df
setPop(pH75_genind_ld_filtered) <- ~Gen

# PCA analysis 
pH8_tab          <- tab(pH8_genind, freq = TRUE, NA.method = "mean")
pH8_pca1         <- dudi.pca(pH8_tab, scale = FALSE, scannf = FALSE, nf = 3)
barplot(pH8_pca1$eig[1:25], main = "PCA eigenvalues", col = heat.colors(25))
s.class(pH8_pca1$li, pop(pH8_genind))


pH8_tab_ld_filt  <- tab(pH8_genind_ld_filtered, freq = TRUE, NA.method = "mean")
pH8_ld_filt_pca1 <- dudi.pca(pH8_tab_ld_filt, scale = FALSE, scannf = FALSE, nf = 3)
barplot(pH8_ld_filt_pca1$eig[1:25], main = "PCA eigenvalues", col = heat.colors(25))
s.class(pH8_ld_filt_pca1$li, pop(pH8_genind_ld_filtered))
title("PCA of dataset\naxes 1-2")
add.scatter.eig(pH8_ld_filt_pca1$eig[1:20], 3,1,2)




# PCA analysis 
pH75_tab          <- tab(pH75_genind, freq = TRUE, NA.method = "mean")
pH75_pca1         <- dudi.pca(pH75_tab, scale = FALSE, scannf = FALSE, nf = 3)
barplot(pH75_pca1$eig[1:25], main = "PCA eigenvalues", col = heat.colors(25))
s.class(pH75_pca1$li, pop(pH75_genind))


pH75_tab_ld_filt  <- tab(pH75_genind_ld_filtered, freq = TRUE, NA.method = "mean")
pH75_ld_filt_pca1 <- dudi.pca(pH75_tab_ld_filt, scale = FALSE, scannf = FALSE, nf = 3)
barplot(pH75_ld_filt_pca1$eig[1:25], main = "PCA eigenvalues", col = heat.colors(25))
s.class(pH75_ld_filt_pca1$li, pop(pH75_genind_ld_filtered))
title("PCA of dataset\naxes 1-2")
add.scatter.eig(pH75_ld_filt_pca1$eig[1:20], 3,1,2)


```



```{r load strata}
F1_Juv_strata_df # we will us this for the intersected juvenile data above
F1_F2_strata_df # for the merged data  we already laoded and fomatted this! THIS IS NOT NORMALIZED AND IS MERGED FOR ALL ALLELES!
F3_Juv_strata_df # 147 rows!
```

## build genind
* use adegent - genind class is a dataframe compatible iwth adegent based on usual genetic markers

```{r genid build}
F1F2_Juveniles_genid <- vcfR2genind(F1F2_Juveniles_vcf)
F1_F2_strata_df$pop <- paste0(F1_F2_strata_df$Gen, F1_F2_strata_df$Treatment)
strata(F1F2_Juveniles_genid) <- F1_F2_strata_df
setPop(F1F2_Juveniles_genid) <- ~pop



# F1 Low and moderate interect genes only 
F1_Juveniles_Intersect_genid <- vcfR2genind(F1_Juveniles_Intersect_vcf)
F1_Juv_strata_df$pop <- paste0(F1_Juv_strata_df$Gen, F1_Juv_strata_df$Treatment)
strata(F1_Juveniles_Intersect_genid) <- F1_Juv_strata_df
setPop(F1_Juveniles_Intersect_genid) <- ~pop





# F1 and F2 low nad moderate interect genes only 
F1F2_Juveniles_Intersect_genid <- vcfR2genind(F1F2_Juveniles_Intersect_vcf)
F1_F2_strata_df$pop <- paste0(F1_F2_strata_df$Gen, F1_F2_strata_df$Treatment)
strata(F1F2_Juveniles_Intersect_genid) <- F1_F2_strata_df
setPop(F1F2_Juveniles_Intersect_genid) <- ~pop


# F3 (low mod and high)  interect genes only 
F3_Juveniles_Intersect_genid <- vcfR2genind(F3_Juveniles_Intersect_vcf)
F3_Juv_strata_df$pop <- paste0(F3_Juv_strata_df$Gen, F3_Juv_strata_df$Treatment)
strata(F3_Juveniles_Intersect_genid) <- F3_Juv_strata_df
setPop(F3_Juveniles_Intersect_genid) <- ~Treatment

# All Juvneiles (low mod and high)  interect genes only 
AllJuveniles_Intersect_genid <- vcfR2genind(AllJuveniles_intersect_vcf)
AllJuveniles_strata_df$pop <- paste0(AllJuveniles_strata_df$Gen, AllJuveniles_strata_df$Treatment)
strata(AllJuveniles_Intersect_genid) <- AllJuveniles_strata_df
setPop(AllJuveniles_Intersect_genid) <- ~pop


# F0 - F1 Merged and Intersect
F0F1_Merged_genid <- vcfR2genind(F0F1_merged.vcf) # 81,325 loci
F0F1_merged.strata$pop <- paste0(F0F1_merged.strata$Gen, F0F1_merged.strata$Treatment)
strata(F0F1_Merged_genid) <- F0F1_merged.strata
setPop(F0F1_Merged_genid) <- ~pop



F0F1_Intersect_genid <- vcfR2genind(F0F1_intersect.vcf) # 6,048 loci;
F0F1_merged.strata$pop <- paste0(F0F1_merged.strata$Gen, F0F1_merged.strata$Treatment)
strata(F0F1_Intersect_genid) <- F0F1_merged.strata
setPop(F0F1_Intersect_genid) <- ~pop


F0F1_pH75_Intersect_genid <- vcfR2genind(F0BroodF1JuvenilespH75.vcf) # 6,048 loci;
F0F1_strata_phH75$pop <- paste0(F0F1_strata_phH75$Gen, F0F1_strata_phH75$Treatment)
strata(F0F1_pH75_Intersect_genid) <- F0F1_strata_phH75
setPop(F0F1_pH75_Intersect_genid) <- ~pop


F0F1_pH8_Intersect_genid <- vcfR2genind(F0BroodF1JuvenilespH8.vcf) # 6,048 loci;
F0F1_strata_phH8$pop <- paste0(F0F1_strata_phH8$Gen, F0F1_strata_phH8$Treatment)
strata(F0F1_pH8_Intersect_genid) <- F0F1_strata_phH8
setPop(F0F1_pH8_Intersect_genid) <- ~pop


```


### Discriminant analysis of principle components (DAPC) 

* find.clusters

  - what is this all about? Investigating genetic diversity using multivarate aprpoches 
  relies on synthetic variables build as linear combinations of alleles which relects the gnetic variation 
  amoung the studied individuals - this allows us in turn to study the diversity beween groups of individuals.
  One route is via genetic clusters where a standard multivariate ANOVA model is used 
  
  total variance = (variance between groups) + (variance within groups)
  
  - PCA focuses on the this total variance or global diveristy, overlooking the differences
  between groups (as the sum of between and within!). 
  
  - DAPC optimizes the between group variance while minimizing the within group variance. 
  Synthetic variables as discriminant functions show differenes between groups while minimizing variation within 
  the cluster
  
  - Note: this is from the adagement tutorial - I am curious, does this DAPC and creation of 
  discriminant functions *bias* the the outcome to focus on between-group variance? OR does it simply
  remove the background to easily discern the present, yet shrouded between-group variance from heavy 
  within-group variance? I hope my future self thinks this makes sence.. 
  
  
```{r find clusters genind object}

grp <- find.clusters(F3_Juveniles_Intersect_genid, max.n.clust=40)

# graph of cumulated variance explained by the eigenvalues of the PCA
# apart from computationa; time, no reason to keep a small number of components 
# we can specify to retain PCs 
table(pop(F3_Juveniles_Intersect_genid), grp$grp)



# run the cluster filters onthe intersect data for F1 pH75 and 8
grp_F1_Intersect <- find.clusters(F1_Juveniles_Intersect_genid, max.n.clust=40)
table(pop(F1_Juveniles_Intersect_genid), grp_F1_Intersect$grp)
table.value(table(pop(F1_Juveniles_Intersect_genid), grp_F1_Intersect$grp), 
            col.lab=paste("inf", 1:20),
            row.lab=paste("ori", 1:20))
# DAPC using function dapc transforms data using PCA and runs discriminant analysis on the retained 
# principle compoenents running on the inferred $grp groups
dapc1 <- dapc(F3_Juveniles_Intersect_genid, grp_F1_Intersect$grp)
scatter.dapc(dapc1, grp=F3_Juveniles_Intersect_genid$pop)
?scatter.dapc
scatter(dapc1, posi.da="topleft", bg="white",
        grp=F3_Juveniles_Intersect_genid$pop,
        pch=17:22, cstar=0,  scree.pca=TRUE,
        posi.pca="bottomleft")

scatter.dapc(dapc1,1,1, 
        grp=F3_Juveniles_Intersect_genid$pop,
        bg="white", scree.da=FALSE, legend=TRUE, solid=.4,
        txt.leg=paste("Cluster",1:6))


```


## build genlight 

* use adegent - genlight is a dataframe compatible with adegenet based on genome-wide SNP data
```{r build genlight}
F1F2_Juveniles_genlight <- vcfR2genlight(F1F2_Juveniles_vcf, n.cores = 4)

```


```{r}
# pairwise Fst
## wrapper for pegas pacakage
F1F1_pH8_perloci.fst     <- Fst(as.loci(F0F1_pH8_Intersect_genid, pop=pop))
F1F1_pH75_perloci.fst    <- Fst(as.loci(F0F1_pH75_Intersect_genid, pop=pop))

F1F1_pH8_perloci.fst_NA  <- F1F1_pH8_perloci.fst %>% na.omit()
F1F1_pH75_perloci.fst_NA <- F1F1_pH75_perloci.fst %>% na.omit()

## wrapper for heirfstat
mat.fst <- genet.dist(F1F2_Juveniles_genid) # (hierfstat): implements pairwise FST , which is also a Euclidean distance between populations.
mat.fst
is.euclid(mat.fst)

?genet.dist

# Fst using genet.dist
# “Dch” By default, Cavalli-Sforza and Edwards Chord distance (eqn 6 in the reference) is returned. This distance is used as default since Takezaki & Nei (1996) found that it was the best to retrieve the relation among samples
F1_intersect.fst      <- genet.dist(F1_Juveniles_Intersect_genid) # default is “Dch”
F1_intersect.fstNei87 <- genet.dist(F1_Juveniles_Intersect_genid, method = "Nei87")



F1F2_intersect.fst      <- genet.dist(F1F2_Juveniles_Intersect_genid) # default is “Dch”
F1F2_intersect.fstNei87 <- genet.dist(F1F2_Juveniles_Intersect_genid, method = "Nei87")



F3_intersect.fst      <- genet.dist(F3_Juveniles_Intersect_genid) # default is “Dch”
F3_intersect.fstNei87 <- genet.dist(F3_Juveniles_Intersect_genid, method = "Nei87")



AllJuveniles_intersect.fst      <- genet.dist(AllJuveniles_Intersect_genid) # default is “Dch”
F3_intersect.fstNei87 <- genet.dist(AllJuveniles_Intersect_genid, method = "Nei87")



F0F1_Merged.fst      <- genet.dist(F0F1_Merged_genid) # default is “Dch”
F3_intersect.fstNei87 <- genet.dist(AllJuveniles_Intersect_genid, method = "Nei87")


F0F1_Intersect.fst      <- genet.dist(F0F1_Intersect_genid) # default is “Dch”


```


```{r}

```


```{r}

```


```{r}

```


```{r}

```


```{r}

```

```{r}

```


```{r}

```


```{r}

```

-   F-statistics: a measure of genetic structure developed by Sewall Wright (1969, 1978). Related to statistical analysis of variance (ANOVA)

    -   FST is the proportion of the total genetic variance contained in a subpopulation (the S subscript) relative to the total genetic variance (the T subscript). Values can range from 0 to 1. High FST implies a considerable degree of differentiation among populations.

    -   FIS (inbreeding coefficient) is the proportion of the variance in the subpopulation contained in an individual. High FIS implies a considerable degree of inbreeding.

    -   Related measures: q (theta) of Weir and Cockerham (1984) and GST of Nei (1973, 1978). [See Weir, 1996; Avise, Box 6.3, p. 206].


```{r}
getwd()
my_vcf <- read.vcfR("C:/Users/samuel.gurr/Documents/F1_F2_Juveniles_Merged.vcf.gz")
ld_filtered_vcf <- my_vcf[!is.na(res$loadings[,1]),]

ld_filtered_vcf_LD <- my_vcf[!is.na(res_LD$loadings[,1]),]

```

-   Next, we will convert our vcfR file to a genlight input which can be used by the package adegenet. We will also use the strata and setPop function to enter population level information. Remember, we made the strata_df in the previous tutorial, but we will remake it here just in case.

```{r}
library(adegenet)
library(hierfstat)
mygl <- vcfR2genlight(ld_filtered_vcf_LD,n.cores =8)
strata(mygl) <- strata_df
setPop(mygl) <- ~Treatment

#  glPca (adegenet): implements PCA for genome-wide SNP data stored as genlight
# objects; see dedicated tutorial (genomics).

myglPCA <- glPca(mygl)
```


```{r Fst}
library(adegenet)
# create genind object from vcf file - use the LD object 
ld_genind <- vcfR2genind(my_vcf)
# assign metadata to these loci 
F1_F2_strata_df
strata(ld_genind) <- F1_F2_strata_df
setPop(ld_genind) <- ~Gen

# pairwise Fst
## wrapper for pegas pacakage
Fst(as.loci(ld_genind))
?Fst
## wrapper for heirfstat
mat.fst <- genet.dist(ld_genind) # (hierfstat): implements pairwise FST , which is also a Euclidean distance between populations.
mat.fst
is.euclid(mat.fst)



# PCS analysis 
X <- tab(ld_genind, freq = TRUE, NA.method = "mean")
class(X) # "matrix" "array"
dim(X) # 79 42782
pca1 <- dudi.pca(X, scale = FALSE, scannf = FALSE, nf = 3)
barplot(pca1$eig[1:50], main = "PCA eigenvalues", col = heat.colors(50))
s.class(pca1$li, pop(ld_genind))
title("PCA of F1 Juvenile dataset\naxes 1-2")
add.scatter.eig(pca1$eig[1:20], 3,1,2)

```

Inbreeding 
refers to an excess of homozygosity in a given individual due to the mating of
genetically related parents. This excess of homozygosity is due to the fact that there are
non-negligible chances of inheriting two identical alleles from a recent common ancestor.
Inbreeding can be associated to a loss of fitness leading to ”inbreeding depression”. Typically,
loss of fitness is caused by recessive deleterious alleles which have usually low frequency in
the population, but for which inbred individuals are more likely to be homozygotes.

```{r inbreeding}

# use sepop to separate pops and explore inbreeding

F1_Low <- seppop(ld_genind)$Low

F1_Mod <- seppop(ld_genind)$Moderate

# compute the mean inbreeding for each individual, and plot the resulting distribution
F1_Low_temp <- inbreeding(F1_Low, N=100)
F1_Mod_temp <- inbreeding(F1_Mod, N=100)

# temp is a list of values sampled from the likelihood distribution of each individual; means
# values are obtained for all individuals using sapply

F1_Low_Fbar <- sapply(F1_Low_temp, mean)
hist(F1_Low_Fbar, col="firebrick", main="Average inbreeding in F1 Low")


# We can see that all  individuals  have higher inbreeding (>0.4).
# We can recompute inbreeding for them, asking for the likelihood function to be  returned

which(F1_Low_Fbar>0.4) # all of them
F1_Low_F <- inbreeding(F1_Low, res.type="function")[which(F1_Low_Fbar>0.4)]

plot(F1_Low_F$adapter_trim.201_querydupscoord.bam, 
     main=paste("Inbreeding of individual",
                names(F1_Low_F)),
     xlab="Inbreeding (F)", 
     ylab="Probability density")


```


-   Below, we use the PCA function of adegent to create a PCA.

```{r}
oyster.pca <- glPca(mygl, nf = 3)
```

-   Let’s look at how much variance each PC explains:

```{r}
barplot(100*oyster.pca$eig/sum(oyster.pca$eig),
        col = heat.colors(50), 
        main="PCA Variance Explained")
title(ylab="Percent of variance\nexplained", 
      line = 2)
title(xlab="PC", 
      line = 1)
```

-   We basically have TWO informative PCs here. Let’s plot the individuals on the first two PCs:

```{r}
myCol <- colorplot(oyster.pca$scores,oyster.pca$scores, 
                   transp=TRUE, 
                   cex=4)
abline(h=0,
       v=0, 
       col="grey")
add.scatter.eig(oyster.pca$eig[1:10],2,1,2, 
                posi="bottomright", 
                ratio=.16)
```

-   The plot above is combining information from the first 3 PCs to color dots, and using the first two for x,y coordinates. We can see possibly four groups across all PCs.

-   We can also port this to ggplot and color by population:

```{r}
library(ggplot2)
oyster.pca.df <- as.data.frame(oyster.pca$scores)
oyster.pca.df$pop <- pop(mygl)
cols <-c("#871a1a","#33A02C","#1F78B4")

p <- ggplot(oyster.pca.df, aes(x=PC1, y=PC2, colour=pop)) 
p <- p + geom_point(size=2)
p <- p + stat_ellipse(level = 0.95, size = 1)
p <- p + scale_color_manual(values = cols) 
p <- p + geom_hline(yintercept = 0) 
p <- p + geom_vline(xintercept = 0) 
p <- p + theme_bw()

p
```

# **DAPC**

-   Discriminant analysis of Principal Components (DAPC), allows us to either infer groups or put in groups *a priori*. DAPC seeks to maximize between group variance while minimizing within group variance. I’ll show you how this works both ways below.

### **Finding clusters**

-   'Adegenet' has a built in method for inferring population groupings or clusters. It runs interactively, so this code block below won’t run on it’s own. You need to paste it into the Console.

```{r}
grp <- find.clusters(mygl, max.n.clust=10)
# choose 40 for the number of OCs to retain
```

For the purpose of picking clusters, the more PCs the better. Pick any number greater than 40.

We’d like to see an elbow here, where the lowest point would be the correct number. This doesn’t always work well for subtle structure. I’m going to pick 3 given the results above.

```{r}
grp <- find.clusters(mygl, max.n.clust=10, n.pca =100, n.clust =3)
```

We can make a simple table to see the assignments:

```{r}
table(pop(mygl), grp$grp)
```


<take some time to interpret what this means here!>

## Discriminant Analysis

This is another interactive function. Run in the console.

```{r}
oyster.dapc <- dapc(mygl, grp$grp)
```

**Now this is different than finding clusters. You can overload the analysis with two many PCs.** For this, you want to choose the lowest number that gets you > 60% of the variance. 
  - I would choose 25 or 30.

This is picking the number of discriminant functions. Because we put in four groups, there are only three.(check this if true with our data) Truly, only the first two are informative. Let’s pick 2.

```{r}
oyster.dapc <- dapc(mygl, n.pca = 25, n.da = 2, grp$grp)
```

Now, let’s plot the analysis

```{r}
scatter(oyster.dapc,col=cols,bg="white", solid=1)
```

We can also plot membership probabilities:

```{r}
compoplot(oyster.dapc, 
          posi="bottom",
          txt.leg=paste("Cluster", 1:3), 
          lab="", 
          ncol=4, 
          xlab="individuals")
```

We can use ggplot to make this easier to read and put in the populations for each individual:

```{r}
library(tidyr)
dapc.df <- as.data.frame(oyster.dapc$posterior)
dapc.df$pop <- pop(mygl)
dapc.df$indNames <- rownames(dapc.df)
cols4 <-c("#871a1a","#33A02C","#1F78B4","#FDBF6F" )

dapc.df <- pivot_longer(dapc.df, -c(pop, indNames))
colnames(dapc.df) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")

p <- ggplot(dapc.df, aes(x=Sample, y=Posterior_membership_probability, fill=Assigned_Pop))
p <- p + geom_bar(stat='identity') 
p <- p + scale_fill_manual(values = cols4) 
p <- p + facet_grid(~Original_Pop, scales = "free")
p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
p <- p + ylab("Posterior membership probability")
p
```

## DAPC with original populations

Again, instead of inferring groups, we can use our original populations.

```{r}
oyster.dapc <- dapc(mygl, 
                    n.pca = 25, 
                    n.da = 2)
scatter(oyster.dapc,
        col = cols, 
        cex = 2, 
        legend = TRUE, 
        clabel = F, 
        posi.leg = "bottomleft", 
        scree.pca = TRUE,
        posi.pca = "topleft", 
        cleg = 0.75, 
        posi.da = "topright")
```

Let’s make this better with ggplot

```{r}
dapc.df <- as.data.frame(oyster.dapc$ind.coord)
dapc.df$pop <- pop(mygl)
dapc.df$indNames <- rownames(dapc.df)

p <- ggplot(dapc.df, aes(x=LD1, 
                         y=LD2, 
                         colour=pop)) 
p <- p + geom_point(size=2)
p <- p + scale_color_manual(values = cols) 
p <- p + geom_hline(yintercept = 0) 
p <- p + geom_vline(xintercept = 0) 
p <- p + theme_bw()+ xlab("DA 1") + ylab("DA 2")+ ggtitle("DAPC with populations as groups")

p
```

**Works cited and acknowledgements**
Code for this tutorial was adapted from the following sources:
https://knausb.github.io/vcfR_documentation/
https://grunwaldlab.github.io/Population_Genetics_in_R/gbs_analysis.html
Documentation from the Adegenet R Package